---
title: "Calculating SPAR Indices with REPS"
author: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating SPAR Indices with REPS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(REPS)
data("aritmethic_data")
```

## Introduction

The `calculate_spar()` function is the central entry point in `REPS` for
calculating a **sales price appraisal ratio (SPAR) index**. SPAR methods compare
transaction prices with appraisal values for the same properties. This provides
a price index without fitting a hedonic regression model.

`calculate_spar()` supports three variants:

* **Arithmetic:** divides the arithmetic mean transaction price by the
  arithmetic mean appraisal value.
* **Geometric:** divides the geometric mean transaction price by the geometric
  mean appraisal value.
* **Unweighted:** calculates the transaction-price-to-appraisal ratio for every
  observation and then takes the arithmetic mean of those ratios.

The valid `method` values are therefore:

```r
c("arithmetic", "geometric", "unweighted")
```

One method returns a data frame. A vector of methods returns a named list with
one data frame per method, following the same pattern as
`calculate_hedonic_index()`.

For every method, the ratio in a period is divided by the ratio in the
reference period of its group. A short-term `growth_rate` of 1 therefore means
that the period is at the same level as its reference period.

## Example Data

The package includes `aritmethic_data`, a compact subset of the SPAR development
data. The spelling of the dataset name is intentionally `aritmethic_data`.

```{r example-data}
head(aritmethic_data)
str(aritmethic_data)
```

The columns used in this vignette are:

* `Period`: monthly period in numeric `YYYYMM` format;
* `Appraisal Year`: appraisal reference year;
* `Price`: transaction price;
* `Property Type`: property-type code;
* `Appraisal Value`: appraisal value.

## Input Parameters

The full function call is:

```r
calculate_spar(
  dataset,
  method,
  period_variable,
  dependent_variable,
  appraisal_variable,
  grouping_variables,
  index_type_variable = NULL,
  reference_period_pattern = "01$",
  chained = FALSE,
  base_value = 100,
  month_modulo = 100,
  parallel = FALSE
)
```

The parameters have the following roles:

| Parameter | Required | Meaning |
|:--|:--:|:--|
| `dataset` | Yes | A data frame containing periods, prices, appraisal values, and grouping variables. |
| `method` | Yes | One or more of `"arithmetic"`, `"geometric"`, and `"unweighted"`. |
| `period_variable` | Yes | A single string naming the period column. |
| `dependent_variable` | Yes | A single string naming the transaction price column. |
| `appraisal_variable` | Yes | A single string naming the appraisal value column. |
| `grouping_variables` | Yes | Columns identifying each short-term series. These normally include the appraisal reference year and property type. |
| `index_type_variable` | For chaining | Columns identifying the independent series to be chained. Every such column must also occur in `grouping_variables`. |
| `reference_period_pattern` | No | Regular expression identifying the reference period in each group. The default `"01$"` matches January in `YYYYMM` periods. |
| `chained` | No | Set to `TRUE` to link short-term series. Default: `FALSE`. |
| `base_value` | No | Starting value of each chained index. Default: 100. |
| `month_modulo` | No | Identifies new segments in numeric periods. With `YYYYMM`, the default 100 identifies January because `YYYY01 %% 100 == 1`. |
| `parallel` | No | Allows independent methods to run in parallel when multiple methods are requested. Default: `FALSE`. |

Column arguments must be supplied as **column names in strings**, not as bare R
column objects. For example, use `period_variable = "Period"`.

## Calculating One SPAR Method

The following call calculates the arithmetic SPAR index separately for every
combination of appraisal year and property type:

```{r arithmetic}
spar_arithmetic <- calculate_spar(
  dataset = aritmethic_data,
  method = "arithmetic",
  period_variable = "Period",
  dependent_variable = "Price",
  appraisal_variable = "Appraisal Value",
  grouping_variables = c("Appraisal Year", "Property Type")
)

head(spar_arithmetic)
```

For the arithmetic and geometric methods, the output includes:

* `numerator_numerator`: the aggregated transaction price;
* `numerator_denominator`: the aggregated appraisal value;
* `ratio_numerator`: their ratio in the current period;
* `ratio_denominator`: the ratio in the reference period;
* `growth_rate`: `ratio_numerator / ratio_denominator`.

The unweighted method returns the two ratio columns and `growth_rate`, but not
the separate price and appraisal aggregates.

## Choosing the Grouping Variables

`grouping_variables` defines the short-term series and is therefore a
methodological choice. In the example data, each appraisal year has its own
reference period. Property types should also be calculated separately. The
appropriate grouping is consequently:

```r
grouping_variables = c("Appraisal Year", "Property Type")
```

Do not include a variable merely because it is present in the input data. Add a
variable only when separate SPAR results are required for each of its values.
Every additional grouping variable must have its own matching reference period.

## Reference Periods

The default pattern `"01$"` identifies periods ending in `01`. For numeric or
character periods such as `202101`, this selects January:

```r
reference_period_pattern = "01$"
```

The pattern must match **at most one period in every group**. If it matches more
than one period in a group, `calculate_spar()` stops because the denominator is
ambiguous. A group without a matching reference period is reported in a warning
and omitted from the result.

For quarterly labels such as `"2021Q1"`, a suitable pattern would be:

```r
reference_period_pattern = "Q1$"
```

Quarterly character labels can be used for an unchained calculation. The
current chaining interface requires numeric or numeric-like periods because it
uses `month_modulo` to detect new segments.

## Calculating Multiple Methods

Supply multiple method names as a character vector to compare SPAR variants.
As with `calculate_hedonic_index()`, a multi-method call returns a named list
containing one data frame per method.

```{r multiple-methods}
spar_methods <- calculate_spar(
  dataset = aritmethic_data,
  method = c("arithmetic", "geometric", "unweighted"),
  period_variable = "Period",
  dependent_variable = "Price",
  appraisal_variable = "Appraisal Value",
  grouping_variables = c("Appraisal Year", "Property Type")
)

names(spar_methods)
head(spar_methods$arithmetic)
head(spar_methods$geometric)
head(spar_methods$unweighted)
```

For a large dataset, independent methods may be calculated in parallel:

```r
spar_methods <- calculate_spar(
  dataset = my_data,
  method = c("arithmetic", "geometric", "unweighted"),
  period_variable = "Period",
  dependent_variable = "Price",
  appraisal_variable = "Appraisal Value",
  grouping_variables = c("Appraisal Year", "Property Type"),
  parallel = TRUE
)
```

## Chaining Short-Term SPAR Indices

Set `chained = TRUE` to link the short-term appraisal-year series. The
`index_type_variable` identifies which series must be chained independently. In
this example, each property type receives its own chained index:

```{r chaining}
spar_chained <- calculate_spar(
  dataset = aritmethic_data,
  method = "arithmetic",
  period_variable = "Period",
  dependent_variable = "Price",
  appraisal_variable = "Appraisal Value",
  grouping_variables = c("Appraisal Year", "Property Type"),
  index_type_variable = "Property Type",
  chained = TRUE,
  base_value = 100,
  month_modulo = 100
)

head(
  spar_chained[
    spar_chained[["Property Type"]] == "A",
    c(
      "period",
      "Appraisal Year",
      "Property Type",
      "growth_rate",
      "growth_factor",
      "Index"
    )
  ]
)
```

The additional output columns are:

* `chaining_true_false`: whether the row starts a new short-term segment;
* `growth_factor`: factor applied to the preceding chained index value;
* `Index`: chained index beginning at `base_value`.

At the first observation of a new segment, the short-term index should equal 1
because that observation is its reference period. Within a segment, the growth
factor is the current short-term index divided by the preceding short-term
index.

If more than one row exists for the same period and index-type series, the
function stops rather than joining unrelated rows. In that situation, add the
necessary series identifier to `index_type_variable`. For example:

```r
index_type_variable = c("Property Type", "Region")
```

Both variables must also be present in `grouping_variables`.

## Plotting All Three SPAR Methods

`calculate_spar()` returns standardized `period` and `Index` columns, so its
single- and multi-method results are compatible with `plot_price_index()`.
Because the example data contain several property types for each period, first
select one property-type series. Plotting several grouped rows as one line would
not represent a single price index.

```{r plot-all-spar-methods, fig.width=7, fig.height=5}
spar_plot_data <- aritmethic_data[
  aritmethic_data[["Property Type"]] == "A",
]

spar_plot_results <- calculate_spar(
  dataset = spar_plot_data,
  method = c("arithmetic", "geometric", "unweighted"),
  period_variable = "Period",
  dependent_variable = "Price",
  appraisal_variable = "Appraisal Value",
  grouping_variables = c("Appraisal Year", "Property Type"),
  index_type_variable = "Property Type",
  chained = TRUE
)

plot_price_index(
  spar_plot_results,
  title = "SPAR Method Comparison - Property Type A"
)
```

## Input Validation and Common Problems

Before calculation, `calculate_spar()` checks the following conditions:

* all named columns exist;
* transaction price and appraisal value are numeric;
* observed prices and appraisal values are finite and strictly positive;
* the reference-period regular expression is valid;
* a reference pattern does not select multiple periods in one group;
* chaining options are valid and chain observations are unique.

Missing transaction prices or appraisal values are handled as pairs. If either
value is missing, that observation is excluded from both sides of the relevant
aggregate. This prevents prices and appraisals from being calculated from
different samples.

For reliable results, users should check the input before calculation:

```r
stopifnot(
  is.numeric(my_data$price),
  is.numeric(my_data$appraisal),
  all(my_data$price > 0, na.rm = TRUE),
  all(my_data$appraisal > 0, na.rm = TRUE)
)
```

## Summary

Use `calculate_spar()` when transaction prices and corresponding appraisal
values are available. The essential workflow is:

1. identify the price, appraisal, and period columns;
2. choose grouping variables that define each short-term series;
3. choose a regular expression that selects one reference period per group;
4. select one or more SPAR methods;
5. when chaining, specify the variables that identify independent long-term
   series and use a numeric period format compatible with `month_modulo`.

The `aritmethic_data` example is structured so that all three SPAR methods and
the chaining workflow can be run directly.
