---
title: "Chaining Price Indices (Annual Overlap Method)"
author: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chaining Price Indices (Annual Overlap Method)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(REPS)
data("data_constraxion")
```

# Introduction

While `calculate_hedonic_index()` computes multilateral or direct indices over an entire dataset, long-term pooled models can suffer from structural market changes over time. To account for shifting consumer preferences, indices are often calculated over shorter periods and linked together.

The `calculate_chained_index()` function acts as a supplementary wrapper. It implements the **Annual Overlap Method**, which splits the data by year, calculates a short-term index using the final period of the previous year as the overlap base, and chains them together into a continuous series.

# Example: Chained Fisher Index

You can use `calculate_chained_index()` with any underlying hedonic method supported by the package. Here is an example using the Fisher index.

```{r}
# Prepare data (log-transform numerical variables)
dataset <- data_constraxion
dataset$floor_area <- log(dataset$floor_area)

# Calculate the chained index
chained_fisher <- calculate_chained_index(
  dataset = dataset,
  method = "fisher",
  period_variable = "period",
  dependent_variable = "price",
  numerical_variables = c("floor_area", "dist_trainstation"),
  categorical_variables = c("dummy_large_city", "neighbourhood_code"),
  reference_period = "2015"
)

head(chained_fisher)
```

# Visualizing the Chained Index

Because the output mirrors the standard REPS dataframe structure, it is fully compatible with the package's built-in plotting utility.

```{r}
plot_price_index(chained_fisher)
```